---
title: "RNA-seq"
author: "Lin"
date: "12/5/2019"
output: html_document


## read original data.csv
```{r}
MyData = read.csv(file="ctc_vs_pt_data.csv", header = TRUE, sep = ",",row.names = NULL) 
```

## convert ID
```{r}
require("biomaRt")
mart = useMart("ENSEMBL_MART_ENSEMBL")
mart = useDataset("hsapiens_gene_ensembl", mart)
annotLookup = getBM(
     mart = mart,
     attributes = c("ensembl_gene_id", "gene_biotype", "external_gene_name"),
     filter = "ensembl_gene_id",
     values = MyData$ensgene,
     uniqueRows = TRUE)
data <- data.frame(MyData[match(annotLookup$ensembl_gene_id, MyData$ensgene),],annotLookup$external_gene_name)
write.csv(data,"ctc_vs_pt_data_v1.csv")
```

## Input real data
```{r}
MyData = read.csv(file = "ctc_vs_pt_data_v1.csv", header = TRUE, fill = T,sep = ",",row.names = NULL) 
```

## remove duplicate rows
```{r}
library(dplyr)
Mydata1 = filter(MyData,ensgene != "5_8S_rRNA" & ensgene !="5S_rRNA" & ensgene !="7SK" & ensgene != "ALG1L9P" & ensgene != "BMS1P4" & ensgene != "CYB561D2" & ensgene != "DUXAP8" & ensgene != "GOLGA8M" & ensgene != "ITFG2-AS1" & ensgene != "LINC01238" & ensgene != "LINC01505" & ensgene !="Mar-01" & ensgene != "Mar-02" & ensgene != "MATR3" & ensgene !="Metazoa_SRP" & ensgene != "PINX1" & ensgene != "POLR2J4" & ensgene !="RMRP" & ensgene != "SCARNA4" & ensgene != "SFTA3" & ensgene !="SNORA16A" & ensgene != "SNORA17B" & ensgene != "SNORA50A" & ensgene != "SNORA62" & ensgene != "SNORA63" & ensgene != "SNORA70" & ensgene !="SNORA71" & ensgene != "SNORA72" & ensgene != "SNORA73" & ensgene != "SNORA74" & ensgene != "SNORA75" & ensgene != "SNORD115" & ensgene !="SNORD116" & ensgene != "SNORD18" & ensgene != "SNORD22" & ensgene !="SNORD27" & ensgene != "SNORD30" & ensgene != "SNORD33" & ensgene != "SNORD38B" & ensgene != "SNORD39" & ensgene != "SNORD3D" & ensgene != "SNORD42" & ensgene != "SNORD63" & ensgene != "SNORD81"& ensgene != "TMSB15B" & ensgene != "U1" & ensgene != "U2" & ensgene !="U3" & ensgene != "U4" & ensgene != "U6" & ensgene != "U7" & ensgene != "U8" & ensgene !="Vault" & ensgene != "Y_RNA" )
```

# Rebuild real data
```{r}
rownames(Mydata1) = Mydata1$ensgene
Mydata1 = Mydata1[,-1]
```

## Set the ggplot main theme
```{r}
library(ggplot2)
main_theme = theme( #panel.background = element_blank(),
                    #panel.grid = element_blank(),
                    #axis.line.x = element_line(size = .5, colour = "black"),
                    #axis.line.y = element_line(size= .5, colour = "black"),
                    axis.title.x = element_text(size = 14, color = "black", hjust = 0.5),
                    axis.title.y = element_text(size = 14,color = "black", hjust = 0.5),
                    axis.ticks = element_line(color = "black"),
                    axis.text = element_text(color = "black", size = 12),
                    plot.title = element_text(hjust = 0.5, size = 15),
                    legend.position = "right", # 'none':remove legend
                    legend.background = element_blank(),
                    legend.key = element_blank(),
                    legend.text = element_text(size = 12),
                    text = element_text(family = "sans", size = 15)
                  )
```

## Summary data
```{r}
library(dplyr)
meancounts = Mydata1 %>% 
    mutate(pt = (pt1 + pt2 + pt3 + pt4 + pt5 + pt6)/6) %>% 
    mutate(ctc = (ctc1 + ctc2 + ctc3 + ctc4 + ctc5 + ctc6 + ctc7 + ctc8)/8)  %>% select(ctc,pt)

meancounts %>% summarize(sum(ctc), sum(pt))

meantotalcounts = data.frame(
    value = c(sum(meancounts$ctc),sum(meancounts$pt)),
    group = c('CTC','PT')
)

# plot bar plot
ggplot(meantotalcounts,aes(group,value)) +
geom_bar(aes(fill = group), stat = "identity", position = "dodge", width = 0.8) +
xlab('') +
ylab('') +
labs(title = "A  Mean Count of Each Group") +
main_theme

# plot point plot
ggplot(meancounts, aes(pt, ctc)) + geom_point() + scale_x_log10() + scale_y_log10() + 
xlab('PT') + ylab('CTC') + labs(title = "B  CTC mean VS PT mean") + main_theme
```

## PCA analysis for all data
```{r}
library(ggbiplot)
library(factoextra)

# compute pca
all.pca = prcomp(t(Mydata1[apply(Mydata1, 1, var)!= 0,]), scale. = TRUE)
all.class = factor(rep(c("CTC", "PT"), c(8, 6)))

# visualize eigenvalues
fviz_eig(all.pca) + ggtitle('A  Scree Plot') + main_theme

# plot pca: method 1
fviz_pca_ind(all.pca,
             col.ind = all.class, 
             repel = TRUE,
             legend.title = "Groups"
             ) + 
    ggtitle('A  PCA') + 
    main_theme

# plot pca: method 2
ggbiplot(all.pca, obs.scale = 1, var.scale = 1,
    groups = all.class, var.axes= 0) +
    scale_color_discrete(name = '') +
    theme(legend.direction = 'horizontal', legend.position = 'top') + 
    ggtitle('PCA') + 
    main_theme
```

## Build a colData object
```{r}
coldata = data.frame(condition = c(rep("CTC",8), rep("PT",6)),
                      row.names = colnames(Mydata1))
```

## Install DESeq2
```{r, echo=FALSE}
##    install.packages("BiocManager")
##    BiocManager::install("DESeq2")
library(DESeq2)

dds = DESeqDataSetFromMatrix(countData = Mydata1,
                              colData = coldata,
                              design = ~ condition)

# add potential features
featureData = data.frame(gene=rownames(Mydata1))
mcols(dds) = DataFrame(mcols(dds), featureData)
mcols(dds)

# specify reference condition
dds$condition = relevel(dds$condition, ref = "PT")

# prefiltering out of lowly expressed genes
keep = rowSums(counts(dds)) >= 10
dds = dds[keep,]
```

## DESeq data analysis
```{r}
dds = DESeq(dds)
dds1 = dds
res = results(dds)
```

## count DEGs
```{r}
sum(res$padj < 0.05 & abs(res$log2FoldChange) > 1, na.rm = TRUE)
```

## Output DEGs
```{r}
res$padj[is.na(res$padj)] <- 1

# select those genes with padj < 0.05 and with a abs of log2FoldChane > 1
DEGs = res[res$padj < 0.05 & abs(res$log2FoldChange) > 1,]
# write.csv(DEGs,file = "DEGs.csv")
```

## Heatmap analysis for the DEGs
```{r}
library(pheatmap)
annotation_col = data.frame(
    CellType = factor(rep(c("CTC", "PT"), c(8, 6))))
rownames(annotation_col) = c(paste("ctc", 1:8, sep = ""),paste("pt", 1:6, sep = ""))

# heatmap with log(RPKM + 1)
pheatmap(log(Mydata1[rownames(Test_DEGs),]+1), scale = "row", cluster_col  = FALSE, annotation_col = annotation_col, main
 = 'A Heatmap') 
```

## Make MA plot and Volcano plot
```{r}
res1 = tbl_df(res)
res1 = res1 %>% mutate(significant = padj < 0.05)
row.names(res1) = row.names(res)

# plot MA plot
res1 %>% ggplot(aes(baseMean, log2FoldChange, col = significant)) + geom_point() + scale_x_log10() + ggtitle("MA Plot") + main_theme

# method 1: plot Volcano plot
res1 %>% ggplot(aes(log2FoldChange, -1*log10(padj), col = significant)) + geom_point() +
    geom_vline(xintercept = c(-1,1), color = 'blue', linetype = 2, size = 1) + 
    geom_hline(yintercept = -log10(0.05), color = 'blue', linetype = 2, size = 1) + 
    ggtitle("Volcano Plot") + 
    main_theme

# method 2: plot Volcano plot
res1[which(res1$padj < 0.05 & res1$log2FoldChange >= 1 ),'diff'] = 'up'
res1[which(res1$padj < 0.05 & res1$log2FoldChange <= -1),'diff'] = 'dowm'
res1[!(res1$diff %in% c('up', 'dowm')),'diff'] = 'no'
 
res1 %>% ggplot(aes(x = log2FoldChange, y = -log10(padj))) +
    geom_point(aes(color = diff), size = 1) +
    scale_colour_manual(limits = c('up', 'dowm', 'no'), values = c('blue', 'red', 'gray40'), labels = c('Upregulated Genes','Downregulated OTUs', 'No diff Genes')) +
    geom_vline(xintercept = c(-1,1), color = 'black', linetype = 2, size = 1) + 
    geom_hline(yintercept = -log10(0.05), color = 'black', linetype = 2, size = 1) + 
    labs(x = 'log2 Fold Change', y = '-log10 FDR p-value') + 
    ggtitle("B  Volcano Plot") + 
    main_theme

```

## plot sepecial gene
```{r}
# method 1: plotCounts
plotCounts(dds1, gene = "HBB") 

# method 2: ggplot
all_genes = as.data.frame(t(Mydata1))
HBB = data.frame(
    log10count = log10(all_genes$HBB),
    CellType = factor(rep(c("CTC", "PT"), c(8, 6)))
)
ggplot(HBB,aes(CellType, log10count, colour = CellType)) +
   geom_boxplot(aes(fill = CellType), alpha = 0.2,
               outlier.colour = "red",
               outlier.shape = 2,
               outlier.size = 5,
               coef = 1.5) + 
    ggtitle("C  HBB") + 
    main_theme
```

## PCA analysis for DGEs
```{r}
# method 1:
# pca
degs.pca = prcomp(t(Mydata1[rownames(DEGs),]), scale. = TRUE)
degs.class = factor(rep(c("CTC", "PT"), c(8, 6)))

# pca plot
library(ggbiplot)
ggbiplot(degs.pca, obs.scale = 1, var.scale = 1,
    groups = degs.class, var.axes= 0) +
    scale_color_discrete(name = '') +
    theme(legend.direction = 'horizontal', legend.position = 'top') + ggtitle('PCA for DGEs') + main_theme

# method 2:
rld = rlogTransformation(dds)
plotPCA(rld)
```

## hclust
```{r}
library(cluster)  
library(factoextra)
library(NbClust)
dis = dist(t(Mydata1[rownames(DEGs),]), method = "euclidean")
dis_hc = hclust(d = dis, method = "complete")
fviz_dend(dis_hc, cex = 0.6)
fviz_dend(dis_hc, cex = 0.6)
fviz_dend(dis_hc, k = 2, 
          cex = 0.5, 
          k_colors = c("#2E9FDF", "#00AFBB"),
          color_labels_by_k = TRUE, 
          rect = TRUE,          
          hang = -1)
```

## Use HTSanalyzeR to do functional analysis
```{r}
library(HTSanalyzeR2)
library(org.Hs.eg.db)
# define the gene set collections
GO_MF = GOGeneSets(species = "Hs", ontologies = c("MF"))
GO_BP = GOGeneSets(species = "Hs", ontologies = c("BP"))
PW_KEGG = KeggGeneSets(species = "Hs")
Msig_c5 = MSigDBGeneSets(collection = "C5", species = "Hs")

ListGSC = list(GO_MF = GO_MF, GO_BP = GO_BP, PW_KEGG = PW_KEGG, Msig_c5 = Msig_c5)

# use log2 fold changes as phenotype for GSEA
data4enrich = res$log2FoldChange
names(data4enrich) = rownames(res)

# select upregulated genes for GSOA
hits = rownames(res)[which(res$padj < 0.05 & res$log2FoldChange > 1)]

# initiate an object of GSCA
gsca = GSCA(listOfGeneSetCollections = ListGSC, geneList = data4enrich, hits = hits)

# preprocessing
gsca = preprocess(gsca, species = "Hs", initialIDs = "SYMBOL", keepMultipleMappings = TRUE, duplicateRemoverMethod = "max", orderAbsValue = FALSE)
```

## Perform gsca analysis
```{r}
gsca = analyze(gsca, para = list(pValueCutoff = 0.05, pAdjustMethod = "BH", 
                                nPermutations = 10, minGeneSetSize = 50, exponent = 1), doGSOA = TRUE, doGSEA = TRUE)
```

## Annotate the gene sets ID to make the results more understandable
```{r}
gsca = appendGSTerms(gsca, goGSCs = c("GO_BP", "GO_MF"), keggGSCs = c("PW_KEGG"), 
                      msigdbGSCs = c("Msig_c5"))
```

## Get the top 5 significant results
```{r}
# hypergeometric test result using MF gene sets from GO
head(getResult(gsca)$HyperGeo.results$GO_MF, 5)

# hypergeometric test result using BP gene sets from GO
head(getResult(gsca)$HyperGeo.results$GO_BP, 5)

# GSEA result using KEGG gene sets
head(getResult(gsca)$GSEA.results$PW_KEGG, 5)

# GSEA result using KEGG gene sets
head(getResult(gsca)$GSEA.results$Msig_c5, 5)

# result both significant regarding to adjust pvalues in hypergeometric test and GSEA using 'c5' gene sets from MSigDB
head(getResult(gsca)$Sig.adj.pvals.in.both$Msig_c5, 5)
```

## Better view the GSEA result for a single gene set
```{r}
topGS = getTopGeneSets(gsca, resultName = "GSEA.results",
                        gscs = c("Msig_c5"), ntop = 100)
#topGS
viewGSEA(gsca, gscName = "Msig_c5", gsName = topGS[["Msig_c5"]][86])
```

## View enrichment map of GO and GSEA result 
```{r}
viewEnrichMap(gsca, resultName = "HyperGeo.results",
              gscs = c("PW_KEGG", "GO_MF"),
              allSig = FALSE, gsNameType = "term", ntop = 5)
viewEnrichMap(gsca, resultName = "GSEA.results",
              gscs = c("Msig_c5"),
              allSig = FALSE, gsNameType = "term", ntop = 5)
```

## Enrichment Map with specific gene sets
```{r}
topGS = getTopGeneSets(gsca, resultName = "GSEA.results",
                        gscs = c("Msig_c5"), ntop = 500)

# define specificGeneset 
Msig_geneset = topGS$Msig_c5[c(c(seq(1,5)),c(61,149,285,357,422))]

# match the names with the geneset
specificGeneset = list("Msig_c5" = Msig_geneset)
viewEnrichMap(gsca, resultName = "GSEA.results", gscs=c("Msig_c5"),
              allSig = FALSE, gsNameType = "term",
              ntop = NULL, specificGeneset = specificGeneset)
```

## Enriched subnetwork analysis
```{r}
pv = res$pvalue
names(pv) = rownames(res)
nwa = NWA(pvalues = pv, phenotypes = data4enrich)
nwa1 = preprocess(nwa, species = "Hs", initialIDs = "SYMBOL",
                   keepMultipleMappings = TRUE, duplicateRemoverMethod = "max")
```

## Creating interactome 
```{r}
nwa2 = interactome(nwa1, species = "Hs", link = "https://downloads.thebiogrid.org/Download/BioGRID/Release-Archive/BIOGRID-3.4.162/BIOGRID-ORGANISM-3.4.162.tab2.zip", genetic = FALSE, force = FALSE)
```

## Performe subnetwork analysis
```{r}
nwa3 = analyze(nwa2, fdr = 8e-02, species = "Hs")
```

## View subnetwork
```{r}
viewSubNet(nwa3)
```

```{r}
report(gsca)
```

```{r}
save(gsca, file = "Galaxy.RData")
```
